---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# riskworkflowr

Reproducible spatial risk analysis workflows for areal units and H3 hexagonal grids, including spatial assignment, aggregation, risk metrics, SMR analysis, and cartographic outputs.

## Installation

```{r eval=FALSE}
# install.packages("pak")

pak::pak("GeoRiskExplorer/riskworkflowr")
```

## Why riskworkflowr?

`riskworkflowr` provides a reproducible workflow for:

```text
points/events
→ spatial assignment
→ aggregation/counts
→ risk metrics
→ cartographic outputs
```

The same workflow works for:

- areal units
- administrative boundaries
- custom polygons
- H3 hexagonal grids

## Calculate rates

```{r}
library(riskworkflowr)

data <- data.frame(
  event_count = c(5, 10, 15),
  population = c(1000, 2000, 3000)
)

risk_calc_rate(
  data = data,
  count_col = "event_count",
  denominator_col = "population"
)
```

## Calculate SMR

```{r}
data <- data.frame(
  event_count = c(5, 10, 20),
  population = c(1000, 2000, 3000)
)

risk_calc_smr(
  data = data,
  observed_col = "event_count",
  denominator_col = "population"
)
```

## Spatial risk workflow

```text
point events
→ spatial assignment
→ event counts
→ join back to polygons
→ risk metrics
→ mapping
```

## Package philosophy

`riskworkflowr` is designed as a reproducible spatial risk workflow framework rather than a generic GIS toolbox.

Core principles:

- reproducibility
- defensible spatial assignment
- auditability of joins
- transparent risk metrics
- workflow consistency across areal units and H3 grids

## Example datasets

The package includes synthetic example datasets for demonstration and testing.

No private or operational data are included.