---
title: "Getting Started with riskworkflowr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with riskworkflowr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

`riskworkflowr` provides reproducible workflows for spatial risk analysis, including:

- spatial assignment
- event aggregation
- risk metrics
- SMR analysis
- cartographic outputs

The package is designed to support consistent workflows across:

- administrative boundaries
- custom areal units
- H3 hexagonal grids

## Why riskworkflowr?

Many spatial risk workflows require analysts to repeatedly combine:

- spatial joins
- event aggregation
- comparative risk metrics
- choropleth mapping
- reproducible analytical workflows

`riskworkflowr` aims to provide a consistent framework for these commonly repeated tasks while remaining compatible with the broader R spatial ecosystem.

## Core workflow

```text
point events
→ spatial assignment
→ aggregation/counts
→ risk metrics
→ mapping
```

A core design principle of the package is:

```text
"units are units"
```

The same analytical workflow should operate consistently across:

- administrative boundaries
- custom polygons
- H3 hexagonal grids
- other areal unit systems

## Install package

```{r eval=FALSE}
pak::pak("GeoRiskExplorer/riskworkflowr")
```

## Load package

```{r}
library(riskworkflowr)
```

## Create example data

```{r}
data <- data.frame(
  event_count = c(5, 10, 15),
  population = c(1000, 2000, 3000)
)
```

## Calculate rates

```{r}
risk_calc_rate(
  data = data,
  count_col = "event_count",
  denominator_col = "population"
)
```

## Calculate SMR

```{r}
risk_calc_smr(
  data = data,
  observed_col = "event_count",
  denominator_col = "population"
)
```

## Methodological scope

The package primarily focuses on practical and reproducible workflows for exploratory spatial risk analysis and communication.

The implemented methods should not be interpreted as replacing more advanced epidemiological, spatial statistical, or inferential modelling approaches where such methods are appropriate and supported by the available data.

## Important assumptions

This package implements practical and reproducible spatial risk workflows commonly used in operational and applied analytical contexts.

Alternative epidemiological and spatial statistical approaches may be more appropriate depending on:

- data availability
- denominator quality
- covariate availability
- inferential objectives
- spatial scale
- analytical assumptions

## Next steps

Additional vignettes provide detail on:

- spatial assignment
- aggregation workflows
- SMR analysis
- H3 workflows
- mapping
- methodological considerations