---
title: "Getting Started with riskworkflowr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Overview

`riskworkflowr` provides reproducible workflows for spatial risk analysis, including:

- spatial assignment
- event aggregation
- risk metrics
- SMR analysis
- cartographic outputs

The package is designed to support consistent workflows across:

- administrative boundaries
- custom areal units
- H3 hexagonal grids

## Core workflow

```text
point events
→ spatial assignment
→ aggregation/counts
→ risk metrics
→ mapping
```

## Install package

```{r eval=FALSE}
pak::pak("GeoRiskExplorer/riskworkflowr")
```

## Load package

```{r}
library(riskworkflowr)
```

## Create example data

```{r}
data <- data.frame(
  event_count = c(5, 10, 15),
  population = c(1000, 2000, 3000)
)
```

## Calculate rates

```{r}
risk_calc_rate(
  data = data,
  count_col = "event_count",
  denominator_col = "population"
)
```

## Calculate SMR

```{r}
risk_calc_smr(
  data = data,
  observed_col = "event_count",
  denominator_col = "population"
)
```

## Important assumptions

This package implements practical and reproducible spatial risk workflows commonly used in operational and applied analytical contexts.

Alternative epidemiological and spatial statistical approaches may be more appropriate depending on:

- data availability
- denominator quality
- covariate availability
- inferential objectives
- spatial scale
- analytical assumptions

## Next steps

Additional vignettes provide detail on:

- spatial assignment
- aggregation workflows
- SMR analysis
- H3 workflows
- mapping
- methodological considerations