---
title: "Risk Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(riskworkflowr)
```

## Overview

`riskworkflowr` includes simple helper functions for calculating commonly used spatial risk metrics.

The current metric functions include:

```text
risk_calc_rate()
risk_calc_poisson_probability()
risk_calc_smr()
risk_calc_location_quotient()
```

These functions are intended to support transparent and reproducible analytical workflows rather than replace specialist epidemiological or spatial statistical methods.

## Example data

```{r}
data <- data.frame(
  event_count = c(5, 10, 20),
  population = c(1000, 2000, 3000)
)
```

## Rates

`risk_calc_rate()` calculates a simple rate:

```text
observed events / denominator × multiplier
```

```{r}
risk_calc_rate(
  data = data,
  count_col = "event_count",
  denominator_col = "population"
)
```

## Poisson probability

`risk_calc_poisson_probability()` estimates the probability of one or more events occurring over a defined period using an observed event frequency.

```{r}
poisson_data <- data.frame(
  event_count = c(1, 5, 10),
  years = c(1, 2, 5)
)

risk_calc_poisson_probability(
  data = poisson_data,
  count_col = "event_count",
  period_col = "years",
  output = "both"
)
```

## Standardised mortality ratio style workflows

`risk_calc_smr()` calculates expected counts and standardised mortality ratio style outputs.

Although named using SMR terminology, the same general structure may be useful for other observed-versus-expected spatial risk workflows where an appropriate denominator is available.

```{r}
risk_calc_smr(
  data = data,
  observed_col = "event_count",
  denominator_col = "population"
)
```

## Location quotient style workflows

`risk_calc_location_quotient()` is currently included as a comparative rate helper.

```{r}
risk_calc_location_quotient(
  data = data,
  observed_col = "event_count",
  denominator_col = "population"
)
```

This function is under review because its current behaviour overlaps conceptually with SMR-style comparative rate workflows.

## Distinctive category workflows

Future versions of `riskworkflowr` are planned to support grouped comparative SMR workflows for identifying comparatively distinctive risk categories within spatial units.

This workflow is intended to support analyses such as:

- dominant injury mechanisms
- distinctive incident types
- leading environmental hazards
- category-based comparative risk profiles

The planned workflow will likely use data containing:

```text
unit_id
category
event_count
denominator
```

and return fields such as:

```text
highest_smr_category
highest_smr
highest_event_count
lowest_smr_category
lowest_smr
lowest_event_count
```

This functionality is currently under active design and methodological review.

## Assumptions

Risk metric workflows assume:

- event counts are suitable for the intended analysis
- denominators are appropriate and interpretable
- spatial units are meaningful for the question being asked
- low counts and small denominators are interpreted carefully
- outputs are reviewed alongside context and data quality

## Limitations and pitfalls

Potential limitations include:

- unstable rates in small areas
- misleading interpretation of rare events
- denominator uncertainty
- inconsistent event coding
- ecological interpretation risks
- overinterpretation of exploratory outputs

## Alternative approaches

Depending on the analytical objective and available data, alternative methods may include:

- direct standardisation
- indirect standardisation
- age-sex stratified SMR
- Bayesian smoothing
- empirical Bayes methods
- Poisson or negative binomial regression
- spatial regression
- cluster detection methods
- hierarchical models

These methods may be preferable where sufficient covariate information, stable denominators, or specific inferential objectives exist.

## Future development

Future development priorities include:

- grouped/category SMR workflows
- clearer distinction between SMR and location quotient style calculations
- category dominance outputs for mapping
- optional low-count thresholds
- clearer interpretation flags
- enhanced examples and visual outputs