---
title: "Risk Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(riskworkflowr)
```

## Overview

`riskworkflowr` includes simple helper functions for calculating commonly used spatial risk metrics.

The current metric functions include:

```text
risk_calc_rate()
risk_calc_poisson_probability()
risk_calc_smr()
risk_calc_location_quotient()
```

These functions are intended to support transparent and reproducible analytical workflows rather than replace specialist epidemiological or spatial statistical methods.

## Example data

```{r}
data <- data.frame(
  event_count = c(5, 10, 20),
  population = c(1000, 2000, 3000)
)
```

## Rates

`risk_calc_rate()` calculates a simple rate:

```text
observed events / denominator × multiplier
```

```{r}
risk_calc_rate(
  data = data,
  count_col = "event_count",
  denominator_col = "population"
)
```

## Poisson probability

`risk_calc_poisson_probability()` estimates the probability of one or more events occurring over a defined period using an observed event frequency.

```{r}
poisson_data <- data.frame(
  event_count = c(1, 5, 10),
  years = c(1, 2, 5)
)

risk_calc_poisson_probability(
  data = poisson_data,
  count_col = "event_count",
  period_col = "years",
  output = "both"
)
```

## Standardised mortality ratio style workflows

`risk_calc_smr()` calculates expected counts and standardised mortality ratio style outputs.

Although named using SMR terminology, the same general structure may be useful for other observed-versus-expected spatial risk workflows where an appropriate denominator is available.

```{r}
risk_calc_smr(
  data = data,
  observed_col = "event_count",
  denominator_col = "population"
)
```

## Distinctive category workflows

`risk_distinct_category()` supports grouped/category comparative risk profiling.

It is intended for datasets where events are grouped by both spatial unit and category, such as injury mechanism, incident type, hazard class, or activity type.

```{r}
category_data <- data.frame(
  unit_id = c("A", "A", "B", "B"),
  category = c("Falls", "Water", "Falls", "Water"),
  event_count = c(10, 2, 3, 8),
  population = c(1000, 1000, 800, 800)
)

risk_distinct_category(
  data = category_data,
  unit_id_col = "unit_id",
  category_col = "category",
  observed_col = "event_count",
  denominator_col = "population",
  min_count = 1
)
```

The output identifies the highest comparative category for each unit and, optionally, the lowest comparative category.

This workflow is intended for exploratory spatial risk profiling and should be interpreted carefully where counts are low, denominators are unstable, or categories are inconsistently coded.

## Assumptions

Risk metric workflows assume:

- event counts are suitable for the intended analysis
- denominators are appropriate and interpretable
- spatial units are meaningful for the question being asked
- low counts and small denominators are interpreted carefully
- outputs are reviewed alongside context and data quality

## Limitations and pitfalls

Potential limitations include:

- unstable rates in small areas
- misleading interpretation of rare events
- denominator uncertainty
- inconsistent event coding
- ecological interpretation risks
- overinterpretation of exploratory outputs

## Alternative approaches

Depending on the analytical objective and available data, alternative methods may include:

- direct standardisation
- indirect standardisation
- age-sex stratified SMR
- Bayesian smoothing
- empirical Bayes methods
- Poisson or negative binomial regression
- spatial regression
- cluster detection methods
- hierarchical models

These methods may be preferable where sufficient covariate information, stable denominators, or specific inferential objectives exist.

## Future development

Future development priorities include:

- grouped/category SMR workflows
- clearer distinction between SMR and location quotient style calculations
- category dominance outputs for mapping
- optional low-count thresholds
- clearer interpretation flags
- enhanced examples and visual outputs