---
title: "Risk Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(riskworkflowr)
```

## Overview

Mapping is often the final communication stage of a spatial risk workflow.

In `riskworkflowr`, mapping functions are intended to support:

- exploratory analysis
- operational review
- comparative spatial interpretation
- communication of broad spatial patterns

The package currently focuses on choropleth-style thematic mapping workflows.

## Core workflow

Typical workflows involve:

```text
point events
→ spatial assignment
→ aggregation/counts
→ risk metrics
→ mapping
```

The main mapping helper function is:

```{r eval=FALSE}
map_risk_choropleth()
```

## Supported map types

Current map types include:

```text
count
rate
probability
smr
category
```

These support both continuous and categorical spatial outputs.

## Example workflow

```{r eval=FALSE}
map_risk_choropleth(
  data = analysis_units,
  fill_col = "smr",
  map_type = "smr"
)
```

## Choropleth mapping considerations

Choropleth maps can be useful for communicating broad spatial patterns, but they also introduce important interpretive challenges.

Map outputs are influenced by:

- classification method
- colour palette
- spatial scale
- denominator quality
- spatial unit design
- event rarity
- data smoothing choices

The same underlying data may appear substantially different depending on these decisions.

## Classification methods

Different classification approaches may produce different visual interpretations.

Common approaches include:

- quantile classification
- equal interval classification
- Jenks natural breaks
- custom thresholds
- binary or categorical thresholds

No single classification method is universally correct.

The most appropriate approach depends on:

- the analytical objective
- data distribution
- intended audience
- communication requirements

## Rates versus counts

Counts and rates communicate different information.

### Counts

Count maps may highlight:

- total burden
- operational workload
- absolute event volume

### Rates

Rate maps may highlight:

- relative concentration
- comparative occurrence
- elevated occurrence relative to exposure

Both may be useful, depending on the question being asked.

## SMR-style mapping

SMR-style maps may support exploratory identification of areas with comparatively elevated or reduced observed-versus-expected outcomes.

However, these maps should not be interpreted as:

- causal evidence
- definitive hotspot detection
- proof of operational failure
- precise estimates of individual risk

Interpretation should consider:

- denominator quality
- event rarity
- small number instability
- broader contextual information

## Category mapping

Categorical maps may support communication of:

- dominant risk types
- comparative categories
- grouped classifications
- future distinctive category workflows

Future versions of `riskworkflowr` are planned to support grouped category dominance outputs derived from grouped comparative SMR workflows.

## H3 and hexagonal mapping

The same mapping workflow may be applied to:

- administrative boundaries
- custom polygons
- H3 grids
- hexagonal grids

Different spatial unit systems may produce different visual impressions of the same underlying events.

Analysts should carefully consider:

- resolution
- sparsity
- interpretability
- operational meaning

when mapping fine-scale grids.

## Assumptions

Mapping workflows assume:

- geometry is valid for the processing context
- spatial units are meaningful
- denominators are appropriate
- classification choices are defensible
- outputs are interpreted carefully

## Limitations and pitfalls

Potential issues include:

- misleading classification
- unstable rates
- sparse denominators
- false precision
- ecological interpretation risks
- overinterpretation of visual patterns
- boundary effects
- inconsistent spatial scales

Maps should generally be interpreted as exploratory communication tools rather than definitive analytical conclusions.

## Alternative approaches

Alternative visualisation approaches may include:

- proportional symbols
- density surfaces
- kernel density estimation
- interactive web mapping
- uncertainty mapping
- temporal animation
- small multiple comparisons
- bivariate mapping

The most appropriate method depends on the intended analytical and communication objective.

## Potential future directions

The following areas are currently being explored or considered as possible future enhancements to `riskworkflowr`.

These items are exploratory only and should not be interpreted as guaranteed future functionality.

Potential areas include:

- grouped/category SMR workflows
- distinctive category identification
- richer H3 workflows
- improved uncertainty handling
- enhanced mapping support