---
title: "Risk Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(riskworkflowr)
```

## Overview

`riskworkflowr` provides small, composable functions for calculating commonly used risk metrics.

The current analytical functions include:

```text
risk_calc_rate()
risk_annualise()
risk_poisson_probability()
risk_calc_smr()
risk_distinct_category()
```

These functions are intended to support transparent and reproducible analytical workflows. Each function performs a relatively narrow analytical task so that assumptions and transformations remain explicit.

They are not intended to replace specialist epidemiological or spatial statistical methods where more complex modelling is required.

## Example data

```{r}
risk_data <- data.frame(
  event_count = c(5, 10, 20),
  population = c(1000, 2000, 3000)
)
```

## Rates

`risk_calc_rate()` calculates a simple rate using an observed count and an appropriate denominator:

```text
observed events / denominator × multiplier
```

```{r}
risk_calc_rate(
  data = risk_data,
  count_col = "event_count",
  denominator_col = "population"
)
```

Rates describe event frequency relative to a denominator such as population or exposure. The denominator and multiplier should be selected to suit the analytical question.

## Annualised event frequency

`risk_annualise()` converts an observed event count over a known observation period into an equivalent annual event frequency.

For example, an observed count collected across five years can be expressed as an expected mean annual count:

```{r}
annual_data <- data.frame(
  area = c("A", "B", "C"),
  event_count = c(1, 5, 10),
  years_observed = c(1, 2, 5)
)

risk_annualise(
  data = annual_data,
  count_col = "event_count",
  period_col = "years_observed",
  annualised_col = "lambda_annual"
)
```

Annualisation is a temporal standardisation step. It is distinct from calculating a population or exposure rate.

## Poisson probability

`risk_poisson_probability()` calculates the Poisson probability of observing **at least `k` events**:

\[
P(X \ge k)
\]

where:

- \(X\) is the number of events occurring in the target period,
- \(\lambda\) is the expected mean number of events in that same target period, and
- `k` is the minimum event count of interest.

The function requires lambda to be supplied explicitly. It does not derive lambda from raw counts or observation periods.

When observed counts cover multiple years, `risk_annualise()` can be used first to create an annual lambda.

```{r}
poisson_data <- data.frame(
  area = c("A", "B", "C"),
  event_count = c(1, 5, 10),
  years_observed = c(1, 2, 5)
)

poisson_result <- poisson_data |>
  risk_annualise(
    count_col = "event_count",
    period_col = "years_observed",
    annualised_col = "lambda_annual"
  ) |>
  risk_poisson_probability(
    lambda_col = "lambda_annual",
    k = 1,
    probability_col = "prob_ge_1_annual"
  )

poisson_result
```

Because `lambda_annual` represents the expected mean annual event count, `prob_ge_1_annual` represents:

\[
P(X \ge 1)
\]

within a one-year period.

A different threshold can be used without changing lambda:

```{r}
poisson_data |>
  risk_annualise(
    count_col = "event_count",
    period_col = "years_observed",
    annualised_col = "lambda_annual"
  ) |>
  risk_poisson_probability(
    lambda_col = "lambda_annual",
    k = 2,
    probability_col = "prob_ge_2_annual"
  )
```

This calculates the probability of observing **two or more events within one year**.

Probabilities are returned as numeric proportions between 0 and 1. Percentage formatting should be applied later in reporting or presentation workflows.

## Standardised mortality ratio style workflows

`risk_calc_smr()` calculates expected counts and standardised mortality ratio style outputs.

Although named using SMR terminology, the same general structure may be useful for other observed-versus-expected spatial risk workflows where an appropriate denominator and reference rate are available.

```{r}
risk_calc_smr(
  data = risk_data,
  observed_col = "event_count",
  denominator_col = "population"
)
```

SMR-style measures compare observed events with the number expected under a reference rate. Interpretation should consider count stability, denominator quality, and the suitability of the reference population.

## Distinctive category workflows

`risk_distinct_category()` supports comparative risk profiling where events are classified into categories within spatial or other analytical units.

Examples include injury mechanism, incident type, hazard class, activity type, or other event classifications.

```{r}
category_data <- data.frame(
  unit_id = c("A", "A", "B", "B"),
  category = c("Falls", "Water", "Falls", "Water"),
  event_count = c(10, 2, 3, 8),
  population = c(1000, 1000, 800, 800)
)

risk_distinct_category(
  data = category_data,
  unit_id_col = "unit_id",
  category_col = "category",
  observed_col = "event_count",
  denominator_col = "population",
  min_count = 1
)
```

The output identifies the strongest comparative category for each unit and, where requested, the weakest comparative category.

This type of analysis is intended primarily for exploratory risk profiling. Results should be interpreted carefully where counts are low, denominators are unstable, or categories are inconsistently coded.

## Composable analytical workflows

The metric functions are designed to be combined where appropriate rather than treated as a mandatory sequence.

For example:

```text
observed events
      |
      v
risk_annualise()
      |
      v
expected annual event count (lambda)
      |
      v
risk_poisson_probability()
      |
      v
P(X >= k) for one year
```

A population-rate analysis may instead use `risk_calc_rate()` directly, while an observed-versus-expected analysis may use `risk_calc_smr()`.

The appropriate sequence depends on the analytical question.

## Assumptions

Risk metric workflows assume that:

- event counts are suitable for the intended analysis,
- observation periods are correctly defined,
- denominators are appropriate and interpretable,
- spatial or analytical units are meaningful for the question,
- low counts and small denominators are interpreted carefully, and
- outputs are reviewed alongside context and data quality.

For Poisson probability calculations, the supplied lambda and event threshold must refer to the same target period.

## Limitations and pitfalls

Potential limitations include:

- unstable rates in small populations or low-exposure areas,
- misleading interpretation of rare events,
- denominator uncertainty,
- inconsistent event coding,
- inappropriate observation-period definitions,
- ecological interpretation risks,
- failure of Poisson assumptions such as equidispersion, and
- overinterpretation of exploratory outputs.

## Alternative approaches

Depending on the analytical objective and available data, alternative methods may include:

- direct standardisation,
- indirect standardisation,
- age-sex stratified SMR,
- Poisson regression,
- negative binomial regression,
- empirical Bayes methods,
- Bayesian smoothing,
- spatial regression,
- cluster detection methods, and
- hierarchical models.

These methods may be preferable where sufficient covariate information, stable denominators, repeated observations, spatial dependence, overdispersion, or specific inferential objectives are present.

## Future development

Future development may include:

- grouped/category SMR workflows,
- optional low-count handling,
- clearer analytical interpretation outputs,
- additional worked spatial examples, and
- enhanced reporting and visualisation workflows.
